An antibody-escape estimator for mutations to the SARS-CoV-2 receptor-binding domain

### 2022-03-07 paper/reviewer_response.tex

\response{Thanks for the nice summary of the work.}

While the significance of the work is very high, the manuscript would be improved by addressing the following points.

For every monoclonal antibody, a single escape fraction value is calculated at each site by summarizing the impact of all 20 amino acid substitutions at the site as either the mean impact or the sum of the impacts to produce their escape maps.  When assessing the impact of a new mutation on binding escape, any mutation at that site appears to be treated equivalently.  An alternative approach would be to just use the specific escape fraction for the specific amino acid substitution at that site.  Thus, E484K would be treated differently from E484Q and E484A.  Indeed, one would expect that some substitutions would be more or less tolerated in terms of antibody binding.  The authors may have already evaluated this approach.  At a minimum, this point should be discussed in the manuscript.  But it might also be worthwhile to compare the performance of this more specific approach to the more generic approach implemented here.

\response{This is a good question, and we have elaborated on this point at both the beginning of the Results (to explain our approach) and in the Discussion (to mention this as a caveat).
The major reason is that in practice mutation-level measurements tend to be a bit noisier in our data due (for instance) to specific mutations at a site sometimes being at low abundance in the deep mutational scanning libraries.
Averaging over mutations at a site decreases the effects of this noise.
For this reason, we think that the approximation of assuming mutations at a site have similar effects (which is often roughly true) is less bad than the decreased accuracy of the mutation-level compared to site-level measurements.
A secondary reason is that using site-level measurements means that the calculations no longer have to be referred to a specific RBD sequence, which is useful if the immunodominance hierarchy remains the same even as the virus evolves and people are exposed to new variants (for instance, making antibodies against A484 rather than E484 after infection with Omicron).
Overall, the new text should make clear that using the site-level measurements is an approximation for the reason the reviewer points out, but why we think it is overall a useful approximation.}

### 2022-03-07 paper/paper.tex
explain site vs mutation level info

\subsection{Combining monoclonal antibody escape maps reveals correlated and independent viral antigenic mutations}
A deep mutational scanning experiment can measure how all single amino-acid mutations to the SARS-CoV-2 RBD affect binding by a monoclonal antibody~\citep{greaney2021complete}.
This mutation-level information can be summarized for each RBD site, such as by taking the mean or sum of mutation-level effects at a site.
Here we will work with these site-level escape maps.

@@

There are caveats that should be kept in mind when using the escape estimator.
First, the estimator only considers sites in the RBD, and ignores mutations to other regions of spike.
Second, the estimator assumes the neutralizing activity of human polyclonal serum is represented by an equipotent mix of the monoclonal antibodies that happen to have been previously characterized by deep mutational scanning.
Third, the estimator simply averages site-level escape measurements across antibodies, and does not yet implement a real biophysical model of the combined activity of multiple antibodies~\citep{einav2020two}.
Finally, and in our minds most significantly, the estimator estimates the impact of mutations in reference to antibodies targeted to the early Wuhan-Hu-1 RBD---an approach that is currently reasonable, but will become problematic as human exposure and vaccination histories diversify in the years to come (see last paragraph).

>>>>

\subsection{Combining monoclonal antibody escape maps reveals correlated and independent viral antigenic mutations}
A deep mutational scanning experiment can measure how all single amino-acid mutations to the SARS-CoV-2 RBD affect binding by a monoclonal antibody~\citep{greaney2021complete}.
This mutation-level information can be summarized for each RBD site, such as by taking the mean or sum of mutation-level effects at a site.
Here we will work with these site-level escape maps.
We use site-level information for two reasons: mutation-level measurements tend to be more noisy and averaging them for a site decreases this noise, and using site-level information makes the approach independent of the particular wildtype amino acid at a site (which is useful if we want to keep using the calculator as the RBD evolves).
However, we note that site-level information ignores the possibility of different mutations at a site having different effects, and so mutation-level approaches could also become useful as the quality of experimental data improve.

@@

There are caveats that should be kept in mind when using the escape estimator.
First, the estimator only considers sites in the RBD, and ignores mutations to other regions of spike.
Second, the estimator assumes the neutralizing activity of human polyclonal serum is represented by an equipotent mix of the monoclonal antibodies that happen to have been previously characterized by deep mutational scanning.
Third, the estimator simply averages site-level escape measurements across antibodies, and ignores differences in the effects of different amino-acid mutations at the same site.
Fourth, the estimator does not yet implement a real biophysical model of the combined activity of multiple antibodies~\citep{einav2020two}.
Finally, and in our minds most significantly, the estimator estimates the impact of mutations in reference to antibodies targeted to the early Wuhan-Hu-1 RBD---an approach that is currently reasonable, but will become problematic as human exposure and vaccination histories diversify in the years to come (see last paragraph).
